\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-09-16 chapters/bci.tex
BCI chapter

\todo{chapters don't nicely follow imrad structure. Is this  problem?}
\todo{concern: is a bit in between formula 1 and formula 2}
\section{Rationale}
\section{Overview \& state-of-the-art}
different recording technologies and level of invasiveness + infographic
active/reactive/passive + infographic
\section{Visual Event-Related Potential based BCIs}

>>>>

\todo{Shortly describe challenges from the patient point of view, related to
this thesis}

\section{A direct interface to the brain}

Brain-computer interfaces (BCIs) read your mind.
BCIs are an entertaining subject with many futuristic and more established
applications.
Devices processing direct inputs from the central nervous system can be
useful for rehabilitation, medical diagnosis or treatment.
They have always been especially promising as assistive technology for
paralyzed patients limited in their communication ability. \todo{expand more on this last
sentence, put patients first in introduction}
Furthermore, they bring a new paradigm to human-computer interaction,
especially when paired with virtual or augmented reality.

Simply put, a BCI records the user's brain activity, then extracts some
relevant output from this brain activity and couples this output to a function
of a device.
Optionally, the user can then observe the action of the device and adapt their
behaviour or brain activity accordingly, closing the loop
\footnote{
  \textcite{BCISociety2024} has recently formalized this into the
  following definition of a BCI:
  \it``A brain-computer interface is a system that measures brain activity and
  converts it in (nearly) real-time into functionally useful outputs to replace,
  restore, enhance, supplement, and/or improve the natural outputs of the brain,
  thereby changing the ongoing interactions between the brain and its external or
  internal environments. It may additionally modify brain activity using targeted
  delivery of stimuli to create functionally useful inputs to the
  brain.''
}.

\fullpagefig{figures/bci/bci_loop.pdf}{The Brain-Computer Interfacing loop.}{fig:bci-loop}

Let's break this definition down and focus on the separatate representing what is happening in the brain
is the neuronal \textit{action potential}.
The brain consists 86 billion neurons, which are highly interconnected cells that are
the smallest units of information processing.
The action potential is an electrical pulse occurring when a neuron receives
input from another neuron and is activated.\todo{kort door de bocht}
A neuron's action potential, or the combined action potentials of groups of
synchronized neurons, thus generate an electrical field in and near the brain,
which can be measured using electrodes.
Fluctuations in the brain's electrical potential are a much faster and more
informative signal for our purposes.
\todo{lower part to section}
While a myriad of other techniques can be used to measure functional brain
activity, such as functional ultrasound imaging (fUS) or magnetoencephalograpy
(MEG), we'll limit ourselves in this work to measuring the electrical field.
Section~\ref{sec:bci-recording} gives an overview of current recording
technologies.\todo{make shorter, move part to section}

\todo{paradigm}
The brain waves form only one part of the interaction scheme presented
in Figure~\ref{fig:bci-loop}.
Most importantly, wee need to know what exactly to look for in this signal.
For this, the user needs to perform a task which evokes specific activity of
which it is known that it can be retrieved from the brain signal.
For example, the imagined movement of an arm is detectable f

Next, this brain signal needs to be interpreted \emph{in function of} the
task we want our BCI to perform.\todo{feature extraction}
The electrical activity of the brain is relatively weak compared to that of the
environment around us, containing electronic devices and other sources of
interference.
Furthermore, the brain is continuously processing information and carrying out
`background tasks', all of which generate their own electrical signals.
On the other hand, the brain activity we are interested in exploiting often
originates from a specific functional system within the brain, and its activity
might not be easily discerned from all other neural and environmental activity,
akin to trying to listen to a single speaker in a noise environment where
everyone is shouting over each other.
Some obvious sources of interference can be easily filtered out, but to obtain
a good measure reflecting only the activity that is relevant for a given BCI
problem.
This problem is called \emph{decoding} and is often solved through supervised
machine learning.
Section~\label{sec:bci-decoding} takes a closer look at some common and
state-of-the art techniques.
A machine learning algorithm can use some training data for which the performed
tasks are known, and learn to recognize these tasks in new data.
The training data can be either obtained from the BCI user themselves or from
other users.
In the first case, the user would perform a short calibration session before
using the BCI, where they are instructed to perform a known task.
This calibration session can be eliminated by training the decoder on
preexisting labeled data from other sessions and users, but this is harder due
to the variability between subjects and sessions.\todo{make shorter, move part
to section}

Finally, the loop can be closed by coupling this output can be coupled to a
device or actuator.
There are two aspects to this.
On the one hand, the BCI gains its function by allowing the user to interact
with their environment.
On the other hand, the actions performed by the BCI can themselves influence
the user's brain activity, creating an adaptive system.
A direct form of closed-loop BCI is the use of a neurostimulator.
Examples of this are deep brain stimulation to mitigate the symptoms of
Parkinson's disease, or the cochlear implant as an auditive prosthesis.
More indirectly modulation can be achieved by sensory stimulation.
This can involve specific, paradigm-related sensory input, like haptic feedback when a
movement BCI detects an intended action, or simply presenting selected actions
back to the user.
The user's brain will then adapt through reinforcement learning, causing
changes in behavior or strategy.
Due to neuroplasticity, the brain's ability to adapt, this can have a positive
impact on BCI performance and opens the pathway to rehabilitation applications.

\todo{check excercise session presentation}

\section{Recording technologies}
\label{sec:bci-recording}
different recording technologies and level of invasiveness + infographic
cost, ease of use
\section{BCI Paradigms}
\label{sec:bci-paradigms}
\fullpagefig{figures/bci/paradigms.eps}{caption}{fig:bci-paradigms}

\section{Decoders}
\label{sec:bci-decoders}
evaluated on
information transfer rate,
roc auc,
accuracy

Calibration, transfer learning, riemannian geometry, (multi)-linear techniques,
paper fabien lotte

active/reactive/passive + infographic

\section{A case study: the visual oddball ERP paradigm}
\subsection{stimulation}
\subsection{The event-related potential}
\todo{figure of P3 matrix}
P300 and other components
\todo{figure of ERP components}
\subsection{Decoding ERPs}
tLDA, beamforming
Linear decoding techniques

### 2024-09-16 chapters/covert_align.tex
BCI chapter

\section{Data collection \& preprocessing}
\section{Results}
\section{Discussion}
\todo{check chapters paper}

\subsubsection{Riemannian Geometry}
We will compare WCBLE, CBLE and tLDA to Riemannian Geometry
approaches that rely on spatial covariance as features.
Together with tLDA, Riemannian Geometry generally reaches state-of-the-art decoding
performance~\cite{Lotte2018}.
We implemented two Riemannian Geometry pipelines.
The first one estimates shrunk covariances from the ERPs filtered with 6 XDAWN
filters, projects these covariances to a tangent space and classifies the
result using $L2$-regularized logistic regression
(XDAWNCov-TS-LR)~\cite{Cecotti2017}.
Secondly, we adopt the pipeline from \cite{Aydarkhanov2020}, since their work shows
favorable performance in the presence of single-trial ERP latency jitter.
Shrunk spatial covariance matrices are estimated from epochs that are
augmented by concatenating the average target and average non-target ERP as
extra channels, projected to tangent space and classified using
$L2$-regularized logistic regression (ERPCov-TS-LR).

\subsection{Experiments}
We evaluate our approach on a
publicly available dataset and one specifically recorded for this study, designed to
probe different modalities of covert VSA.

\subsubsection{CVSA-ERP dataset}
We recorded a dataset to validate our approach.
The Covert Visuospatial Attention ERP (CVSA-ERP) dataset
consists of 15 participants, mean age $26.34\pm3.04$ years.
This study was approved by the Ethics Commission of University Hospital Leuven
(S62547).
\ref{app:dataset} details the stimulation and recording procedure.
Each subject performed different VSA conditions (overt, covert and split),
illustrated in Figure~\ref{fig:interface}.
Using a hexagonal layout interface, similar to the visual Hex-o-Spell proposed
by \cite{Treder2010}, we present six flashing targets (without letters or
symbols) to the participant while the EEG, electrooculogram (EOG), and the
participant's eye gaze using eye tracking were recorded.
The VSA conditions described in the first row of Figure~\ref{fig:interface} are
considered.
\begin{figure*}
  \bigskip
	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:interface}
		\includegraphics[width=\linewidth]{figures/covert_align/figure3a.pdf}
	\end{subfigure}

@@

@@

\subsubsection{Data preprocessing and analysis}
Analysis was performed using Python and the MNE software package (version
1.3.1)~\cite{Gramfort2013}.
All datasets were band-pass filtered between 0.1Hz and 20Hz with a 4th-order Butterworth filter.
Bad channels in the data were automatically detected using the RANSAC
method~\cite{Fischler1981} and rejected.
The recorded EEG was re-referenced
offline to the average of the mastoid electrodes TP9 and TP10.
Next, the EEG signals were corrected for eye movement artifacts using an
Independent Component Analysis (ICA).
Since we have access to EOG data for the CVSA-ERP dataset components correlating
significantly with the EOG were rejected.
For the BNCI2014-009 dataset, ICA components were manually rejected.
Finally, the EEG signal is divided into epochs ranging from 100ms before stimulus onset to 700ms after stimulus onset and down-sampled to 128Hz.
In both datasets, only 16 channels were kept for
analysis (Fz, FCz, Cz, CPz, Pz, Oz, F3, F4, C3, C4, CP3, CP4, P3, P4, PO7 and
PO8).

>>>>

\section{Introduction}

frenzel
\todo{part about split attention etc from chapter on covert alignment to here}

\section{Materials \& methods}

\subsection{Data collection}
We evaluate our approach on a
publicly available dataset and one specifically recorded for this study, designed to
probe different modalities of covert VSA.

We recorded a dataset to validate our approach.
The Covert Visuospatial Attention ERP (CVSA-ERP) dataset
consists of 15 participants, mean age $26.34\pm3.04$ years.
This study was approved by the Ethics Commission of University Hospital Leuven
(S62547).
\ref{app:dataset} details the stimulation and recording procedure.
Each subject performed different VSA conditions (overt, covert and split),
illustrated in Figure~\ref{fig:interface}.
Using a hexagonal layout interface, similar to the visual Hex-o-Spell proposed
by \cite{Treder2010}, we present six flashing targets (without letters or
symbols) to the participant while the EEG, electrooculogram (EOG), and the
participant's eye gaze using eye tracking were recorded.
The VSA conditions described in the first row of Figure~\ref{fig:interface} are
considered.
\begin{figure*}
  \bigskip
	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:interface}
		\includegraphics[width=\linewidth]{figures/covert_align/figure3a.pdf}
	\end{subfigure}

@@

\subsubsection{CVSA-ERP dataset}
EEG for the CVSA-ERP dataset was recorded using a SynAmps RT amplifier
(Compumedics Neuroscan, Australia) at 2048Hz and 62 Ag/AgCl active electrodes arranged in the
international 10-10 layout fitted to a standard electrode cap (EASYCAP GmbH,
Germany), with electrodes located at AFz and FCz as ground and reference respectively.
Using electrolyte gel, electrode impedances were brought below 5k$\Omega$.
Electrodes TP9 and TP10, used for off-line re-referencing, were directly
attached to the skin using stickers for better contact.
The power line frequency in Belgium is 50 Hz.
Participant's eye gaze was registered using an EyeLink 1000 Plus eye tracker (SR Research,
Canada) in non-fixation mode.

Participants signed the informed consent form and were seated at a distance of
60 cm before a CRT-emulating monitor (VPixx
Technologies, Canada) operating at a refresh rate of 120Hz, displaying 6
circular white targets with a diameter of 4.15° visual angle and laid out on a hexagon
with a radius of 12.28° of visual angle centered on the midpoint of the screen,
conforming to the interface proposed by \cite{Treder2010} (Figure~\ref{fig:layout_targets}).
A hexagonal layout interface with an empty center and a low number of targets
counteracts target crowding and, as long as the subject’s gaze is within the hexagon of
targets, no other target can be between the subject’s gaze and a covertly
attended target.
Targets are full-contrast white and were intensified by scaling them to a
larger size (5.60° of visual angle, Figure~\ref{fig:layout_intense}) instead of changing the contrast to avoid Troxler-fading\footnote{The optical illusion of disappearing unchanging stimuli
experienced when visually fixating~\cite{Troxler1804}.}~\cite{Treder2010} in the
peripheral visual field.
Stimuli were presented using PsychopPy (version 2023.1.3)~\cite{Peirce2019}.

The participant was instructed to press the space bar when ready for a block
of stimulations.
Then, one target was indicated as the cue and the participant
was instructed to count the number of intensifications of the cued target
during the following block of stimulations.
After pressing the space bar again, a blue crosshair appeared, and the subject
was instructed to fixate their gaze on the blue crosshair for the duration of
the stimulation block (Figure~\ref{fig:layout_cross_overt} and Figure~\ref{fig:layout_cross_covert}).
The position of this crosshair determined the VSA condition for this trial:
overt VSA when the crosshair was at the same location as the cued target,
covert VSA when the crosshair appeared in the center of the screen, and split
VSA when the crosshair appeared on a different target than the cued one.
After pressing the space bar again and a delay of 5 seconds, the stimulation block
starts.
All targets were intensified for a duration of 100 ms, in pseudorandom order.
The inter-stimulus-interval (ISI), the time between the onsets of subsequent intensifications,
was variable and consisted of a fixed 300ms interval (of which 100ms with an intensified target onscreen)
with 200ms uniform jitter added, resulting in an ISI
between 200 and 400 ms.
Inter-stimulus intervals were jittered to counteract steady-state effects and residue in averaging. A
longer inter-stimulus interval will increase component amplitude and aid in counteracting temporal
autocorrelation for a higher statistical test precision.
In a block of stimulations, each target was intensified a pseudorandom number of times between 10 and 15.
This led to stimulation blocks with an average duration of 26.25 seconds. After a block of stimulations, an
input prompt appeared to enter the mentally counted number of intensifications.
After inputting this number, the subject was allowed to pause until pressing the space bar again.
In total, six blocks were presented for overt VSA, six blocks for covert VSA, 12 blocks for
split ($d=1$) VSA, 12 blocks for split ($d=2$) VSA and 6 blocks for split
($d=3$) VSA, covering all possible combinations of VSA conditions, cued targets and
crosshair locations.
The experiment started with five non-recorded practice stimulation blocks, one for
each of the 5 VSA conditions.
During these practice blocks, the participant received feedback about their gaze
position and counting accuracy.
Counting the instructions and the participant's response to the
input prompts, a block lasted about 30
seconds. In sum, the experiment featured approximately 45 minutes of
stimulation time.
After blocks 14 and 28, the participant was allowed to take a longer break.
Including these longer breaks, the experiment lasted approximately one hour.
\subsubsection{Riemannian Geometry}
We will compare WCBLE, CBLE and tLDA to Riemannian Geometry
approaches that rely on spatial covariance as features.
Together with tLDA, Riemannian Geometry generally reaches state-of-the-art decoding
performance~\cite{Lotte2018}.
We implemented two Riemannian Geometry pipelines.
The first one estimates shrunk covariances from the ERPs filtered with 6 XDAWN
filters, projects these covariances to a tangent space and classifies the
result using $L2$-regularized logistic regression
(XDAWNCov-TS-LR)~\cite{Cecotti2017}.
Secondly, we adopt the pipeline from \cite{Aydarkhanov2020}, since their work shows
favorable performance in the presence of single-trial ERP latency jitter.
Shrunk spatial covariance matrices are estimated from epochs that are
augmented by concatenating the average target and average non-target ERP as
extra channels, projected to tangent space and classified using
$L2$-regularized logistic regression (ERPCov-TS-LR).

@@

\subsection{Data preprocessing and analysis}
Analysis was performed using Python and the MNE software package (version
1.3.1)~\cite{Gramfort2013}.
All datasets were band-pass filtered between 0.1Hz and 20Hz with a 4th-order Butterworth filter.
Bad channels in the data were automatically detected using the RANSAC
method~\cite{Fischler1981} and rejected.
The recorded EEG was re-referenced
offline to the average of the mastoid electrodes TP9 and TP10.
Next, the EEG signals were corrected for eye movement artifacts using an
Independent Component Analysis (ICA).
Since we have access to EOG data for the CVSA-ERP dataset components correlating
significantly with the EOG were rejected.
For the BNCI2014-009 dataset, ICA components were manually rejected.
Finally, the EEG signal is divided into epochs ranging from 100ms before stimulus onset to 700ms after stimulus onset and down-sampled to 128Hz.
In both datasets, only 16 channels were kept for
analysis (Fz, FCz, Cz, CPz, Pz, Oz, F3, F4, C3, C4, CP3, CP4, P3, P4, PO7 and
PO8).

### 2024-09-16 chapters/cv.tex
Add figures
BCI chapter

\newenvironment{cvlist}{
	\begin{description}[font=\normalfont]\setlength\itemsep{0em}}
		{\end{description}}

@@

% Curriculum vitae =============================================================
\chapter{Curriculum vitae}
\section*{About me}

I am a Ph.D. candidate\todo{change} at KU Leuven Laboratory for Neuro- and Psychophysiology
and the University of Lille Research Center for Informatics, Signal Processing,
and Control Science. My research is centered on developing brain-computer
interface (BCI) technologies with a primary focus on assisting individuals with
severe disabilities, especially those with motor and eye movement impairments.
My goal is to enhance their quality of life by creating accurate and
user-friendly BCIs. Combining my expertise in machine learning, signal
processing, and user interface design with insights from neurophysiology, my
PhD centers around the development of an event-related potential BCI system
using EEG for real-time communication. I strongly believe in the potential of
innovative neurotechnology to empower individuals and patients, enabling them
to lead more independent and fulfilling lives.

\paragraph{Keywords:} brain-computer interfaces, event-related
potentials, multilinear classification, covert visual attention

@@

\subsection*{Awards \& funding}
\begin{cvlist}
	\cvitem{Clinical Trial Sponsorship Program}{Zeto, 2024 \\
		Our research project titled ``A gaze-independent Visual Brain-Computer
		Interface for use by patients with limited or no eye control'' won a spot in
		the Zeto Clinical Trial Sponsorship Program, granting us the resources to conduct research
		with the Zeto EEG headset.
	}
	\cvitem{Student BCI competition 4th place worldwide}{NeuroTechX, 2022 \\
		Our NeuroTechLeuven BrainBrowsR project, a plug-and-play BCI system that lets you control
		social media applications  trough SSVEP-BCI was awarded 90 out of 100 points by
		the jury of the NeuroTechX international student club competition.
	}
	\cvitem{Global Ph.D. Partnership Grant}{KU Leuven and University, 2021--2024}
\end{cvlist}

@@

\subsubsection{Master theses and internships}
\begin{cvlist}
	\cvitem{Eye-tracker and ERP data fusion for gaze-independent visual
		BCI}{\ \\ Juliette  Meunier, University of Lille, 2024}
	\cvitem{LDA in	the combined
		space-time-frequency domain\\ for BCI decoding}{\ \\Lunkyadi Sucipto, KU Leuven,
		2023--2023}
	\cvitem{Tackling the Midas Touch problem in eye tracking with a
		BCI}{\ \\Reniflal Ebenezer Sundaralal, KU Leuven, 2023--2023}
	\cvitem{An EEG connectivity analysis of Alzheimer’s Disease
		and Frontotemporal Dementia}{Zoe Barinaga, KU Leuven, 2023--2024}
	\cvitem{Single-trial ERP latencies as a predictor for the mode
		of visual attention}{\ \\Yildiz Dilara Parry, KU Leuven, 2023}
	\cvitem{A hybrid P300-gaze BCI alternative for navigating virtual
		spaces}{\ \\Gijs Claes, KU Leuven, 2021--2022}
\end{cvlist}

@@

\subsection*{Professional and volunteering Experience}
\begin{cvlist}
	\cvitem{Ambulance EMT}{Red Cross, FAST	vzw, 2022-ongoing}
	\cvitem{Python developer}{Mindspeller, 2019
		Python Flask developer in a KU Leuven spin-off that provides neuromarketing services based on
		scientifically validated neuroscience and AI research.}
	\cvitem{Freelance full-stack web developer}{Self-employed,
		2014--2020}
\end{cvlist}

@@

\subsection*{Languages}
\begin{cvlist}
	\cvitem{Dutch}{Native}
	\cvitem{English}{Fully proficient}
	\cvitem{French}{Advanced}
	\cvitem{German}{Intermediate}
\end{cvlist}

>>>>

\newenvironment{cvlist}{
	\begin{description}[font=\normalfont]%\setlength\itemsep{0em}
}{
  \end{description}
}

@@

% Curriculum vitae =============================================================
\chapter{Curriculum vit\ae}

\subsection*{About me}
I am a Ph.D. candidate\todo{change} at KU Leuven Laboratory for Neuro- and Psychophysiology
and the University of Lille Research Center for Informatics, Signal Processing,
and Control Science. My research is centered on developing brain-computer
interface (BCI) technologies with a primary focus on assisting individuals with
severe disabilities, especially those with motor and eye movement impairments.
My goal is to enhance their quality of life by creating accurate and
user-friendly BCIs. Combining my expertise in machine learning, signal
processing, and user interface design with insights from neurophysiology, my
PhD centers around the development of an event-related potential BCI system
using EEG for real-time communication. I strongly believe in the potential of
innovative neurotechnology to empower individuals and patients, enabling them
to lead more independent and fulfilling lives.

\subsection*{Keywords}
brain-computer interfaces, event-related
potentials, (multi)linear classification, covert visual attention

@@

\subsection*{Awards \& funding}
\begin{cvlist}
	\cvitem{Clinical Trial Sponsorship Program}{Zeto, 2024 \\
		Our research project titled ``A gaze-independent Visual Brain-Computer
		Interface for use by patients with limited or no eye control'' won a spot in
		the Zeto Clinical Trial Sponsorship Program, granting us the resources to conduct research
		with the Zeto EEG headset.
	}
	\cvitem{Student BCI competition 4th place worldwide}{NeuroTechX, 2022 \\
		Our NeuroTechLeuven BrainBrowsR project, a plug-and-play BCI system that lets you control
		social media applications  trough SSVEP-BCI was awarded 90 out of 100 points by
		the jury of the NeuroTechX international student club competition.
	}
	\cvitem{Global Ph.D. Partnership Grant}{KU Leuven and University of Lille, 2021--2024}
\end{cvlist}

@@

\subsubsection{Master theses and internships supervised}
\begin{cvlist}
	\cvitem{Eye-tracker and ERP data fusion for gaze-independent visual
		BCI}{\ \\ Juliette  Meunier, University of Lille, 2024}
	\cvitem{LDA in	the combined
		space-time-frequency domain\\ for BCI decoding}{\ \\Lunkyadi Sucipto, KU Leuven,
		2023--2023}
	\cvitem{Tackling the Midas Touch problem in eye tracking with a
		BCI}{\ \\Reniflal Ebenezer Sundaralal, KU Leuven, 2023--2023}
	\cvitem{An EEG connectivity analysis of Alzheimer’s Disease
		and Frontotemporal Dementia}{Zoe Barinaga, KU Leuven, 2023--2024}
	\cvitem{Single-trial ERP latencies as a predictor for the mode
		of visual attention}{\ \\Yildiz Dilara Parry, KU Leuven, 2023}
	\cvitem{A hybrid P300-gaze BCI alternative for navigating virtual
		spaces}{\ \\Gijs Claes, KU Leuven, 2021--2022}
\end{cvlist}

@@

\subsection*{Professional experience}
\begin{cvlist}
	\cvitem{Ambulance EMT}{Red Cross, FAST	vzw, 2022-ongoing}
	\cvitem{Python developer}{Mindspeller, 2019
		Python Flask developer in a KU Leuven spin-off that provides neuromarketing services based on
		scientifically validated neuroscience and AI research.}
	\cvitem{Freelance full-stack web developer}{self-employed,
		2014--2020}
\end{cvlist}

@@

\subsection*{Languages}
\begin{cvlist}
	\cvitem{Dutch}{native}
	\cvitem{English}{fully proficient}
	\cvitem{French}{advanced}
	\cvitem{German}{intermediate}
\end{cvlist}

### 2024-09-16 chapters/gaze_independence.tex
BCI chapter

\chapter{Gaze-independent BCIs}
\todo{mention partially published}
\chapter{Gaze-independent BCIs}
\section{Oculomotor deficits in Locked-in Syndrome}
\section{Gaze-independent visual BCIs}
\section{Benefits \& drawbacks of covert attention in BCI operation}
\section{Electrophysiological correlates of covert attention}
\section{Compensating for covert attention}
\todo{spatially vs temporally organized interfaces}

Gaze-independent ERP-based BCIs~\cite{Riccio2012} can be realized in three
ways. Firstly, non-visual stimulation paradigms such as auditory and somatosensory paradigms
do not rely on gaze redirection but often result in lower information transfer
rates, increased mental effort and user-dependent variability~\cite{Reichert2020b}.

Secondly, visual stimulation can be optimized, e.g. such that the stimuli are always
present in the field of view either overtly~\cite{Acqualagna2013, Won2018,
Lin2018} or covertly~\cite{Treder2010,Pires2011,Lees2018}. Some noteworthy examples include the GIBS Block Speller~\parencite{Pires2011},
the GeoSpell interface~\parencite{Aloise2012}, and the RSVP
speller~\parencite{Acqualagna2011}.
Non-spatial visual attention (feature attention) can also be exploited, such as
attention to stimulus color, shape or symbol~\cite{Zhang2010,Treder2011,Hwang2015}.
Alternative visual stimulation paradigms can modulate specific ERP components
that are more sensitive to stimulation in the visual periphery~\cite{Schaeff2012,Xu2022}.
These systems typically require users to focus on a central fixation point while
selecting peripheral targets.
These methods, However, can suffer too from reduced information
transfer rates~\cite{Chennu2013}.
However, they still rely to some extent on
eye motor control, often necessitating central gaze fixation.
\todo{advantages
disadvantages rewrite (requires gaze fixation, low itr, not tested with split attention)}

Thirdly, stimuli can be presented in a standard BCI paradigm, but visuospatial attention
can be decoded separately from gaze direction.
Earlier work has made advances in decoding lateralized covert
VSA harnessing the N2pc ERP component~\cite{Thiery2016,Reichert2020b,Wang2022},
decoding covert VSA shifts from spectral content~\cite{Tonin2013}, or hybrid
stimulation paradigms~\cite{Egan2017}.
Since these methods often require a slower stimulation pace than the classical
ERP BCI paradigm, gaze-independent decoding in fast-paced ERP paradigm stimulation
remains underexplored.
\cite{Aloise2012b} aimed to bridge the performance gap between covert and
overt VSA decoding performance.
They compared classical linear and non-linear ERP classifiers on a covert
attention P3 ERP component dataset.
The results revealed no significant performance improvement in covert VSA
decoding for any of the investigated decoders.
\cite{Arico2014} observed higher variability in single-trial P3 peak
latencies relative to stimulus onset during covert VSA compared to overt VSA.
This latency variability contributes to reduced covert VSA decoding performance.
While they proposed an analysis and performance prediction method, they did not provide a
decoding solution.
They suggested that compensating for latency jitter could enhance covert VSA
decoding, but did not actually verify this hypothesis directly.
Additionally, \cite{hardiansyah2020single} developed a classifier for
covert VSA ERPs, exploiting single-trial latency features in combination with
amplitude features for classification with a support vector machine.
They demonstrated the positive influence of single-trial ERP component latency
features on covert VSA inference yet did not attempt to correct the amplitude
features for these latencies.
\cite{Frenzel2011} introduced a similar protocol to our split VSA
setting.
They showed that it is possible to perform split VSA and that, in this
case, visuospatial attention and gaze direction can separately be decoded using
classical ERP techniques.
To the best of our
knowledge, this is the sole study that investigated split attention in ERP-based BCIs.
However, \cite{Frenzel2011} considered their interface only for the case
where the user actively intends to select both targets determined by the gaze
and the VSA.
In the split VSA setting considered in our work, we rather instruct the participant to ignore
the distractor and only attend the cued target, since we are interested in
decoding the visuospatial attention only.

>>>>

\todo{mention partially published}
\todo{spatially vs temporally organized interfaces}
\todo{talk about covert shifts in spatial attention alpha}

\section{The gaze-dependence problem}

One of the goals of Brain-computer interfacing is establishing a communication
channel that does not rely on speech or
muscular activity~\cite{Naci2012,Chaudhary2016}, which in turn can provide
solutions to paralyzed individuals.
%The quest for performant and affordable solutions is most evident in
%visual BCIs using the electroencephalogram (EEG), where it has led to a gamut
%of increasingly sophisticated decoders, tailored to the needs
%of specific stimulation paradigms and usage contexts.

Traditional visual BCI scenarios require the user to overtly direct their
visuospatial attention (VSA) and gaze toward the screen target they intend to select.
In most settings, screen targets are overlaid with non-overlapping, transient
stimuli that evoke event-related EEG potentials (ERPs).
The selected target can be decoded from these ERPs, as is the case for the
oddball paradigm where observing a rare but attended stimulus evokes a P3 ERP
component.
However, a critical challenge arises when users rely solely or in part on covert
VSA, which involves directing visuospatial attention without corresponding eye gaze.
In these cases, classical solutions often fall short of the widely accepted
80\% target selection accuracy threshold deemed necessary for a comfortable user
experience~\cite{Brunner2010,Frenzel2011,Treder2010,Ron2019,Neeling2019},
calling for alternative, gaze-independent solutions.

In this work, we will use the term \emph{gaze-independent} meaning
\emph{‘dealing explicitly with the fact that a user cannot control their
gaze.'}
In the context of a visual BCI, this means that the user's
visuospatial attention and their gaze do not necessarily coincide.
Gaze-independent paradigms are particularly
promising for individuals with impaired eye motor control, such as those
suffering from specific types or stages of Amyotrophic Lateral Sclerosis (ALS),
Multiple Sclerosis (MS), stroke, or brain stem
stroke.
For these patients, gazing directly at a screen target may be uncomfortable, impractical,
or even impossible.
Hence, assistive devices that rely on eye tracking are often inefficient for
them.
Consequently, while BCIs hold great promise for these individuals, conventional
gaze-dependent BCI solutions do not meet their needs due to the absence of gaze
control. Therefore, the development of decoding strategies that account for covert
VSA becomes crucial in the pursuit of high-performance gaze independent
BCIs.

\section{Incidence of eye motor impairment}
Table~\ref{tab:incidence} reports the relatively high frequencies of
eye motor problems, which can range from involuntary eye movements to control
issues and even to partial or complete ophtalmoplegia.
\begin{table}
	\centering
	\begin{tabular}{@{}|l|rrrrr|r|@{}}
		\hline
		         & ALS  & MS   & Stroke  &
		DMD      & SMA  & LiS                           \\ \hline
		Minor    & 50\% & 31\% & 40-70\% & + & - &      \\
		Severe   & 33\% & 3\%  & +       & - & - & 98\% \\
		Complete & 17\% & -    & +       & - & - & 2\%  \\
		\hline
	\end{tabular}
	\caption{Incidence of eye motor impairment in Brain-Computer Interface target
		populations: (ALS: Amyotrophic Lateral Sclerosis, MS: Multiple Sclerosis,
		DMD: Duchenne's Muscular Dystrophy, SMA: Spinal Muscular Atrophy, LiS:
		Locked-In-Syndrome).}\label{tab:incidence}
\end{table}
\todo{document methodology}
\todo{find more references, fill out}
\todo{
Patients with
Totally locked in/locked in with eye motor issues due to
Brain stem stroke
ALS
MS
Motor and eye motor impairments due to neurotrauma (tetraplegia)
...
}

\todo{define eye motor issues}
\todo{systematic review of prevalences and prevalences of eye motor issues per disorder}
\todo{
cite{Guo2022}
cite{Prasad2010}
cite{Serra2018}
cite{Bogousslavsky1987}
cite{Castelnovo2016}
cite{Polet2020}
}

\section{Gaze-independent BCI paradigms}
\label{sec:gaze-independence-paradigms}
Gaze-independent ERP-based BCIs~\cite{Riccio2012, Aloise2012} can be realized in three
ways. Firstly, non-visual stimulation paradigms such as auditory and somatosensory paradigms
do not rely on gaze redirection but often result in lower information transfer
rates, increased mental effort and user-dependent variability~\cite{Reichert2020b}.
\todo{insert table from riccio in text as numbers, not as table}

Secondly, visual stimulation can be optimized, e.g. such that the stimuli are always
present in the field of view either overtly~\cite{Acqualagna2013, Won2018,
Lin2018} or covertly~\cite{Treder2010,Pires2011,Lees2018}. Some noteworthy examples include the GIBS Block Speller~\parencite{Pires2011},
the GeoSpell interface~\parencite{Aloise2012}, and the RSVP
speller~\parencite{Acqualagna2011}.
Non-spatial visual attention (feature attention) can also be exploited, such as
attention to stimulus color, shape or symbol~\cite{Zhang2010,Treder2011,Hwang2015}.
Alternative visual stimulation paradigms can modulate specific ERP components
that are more sensitive to stimulation in the visual periphery~\cite{Schaeff2012,Xu2022}.
Nevertheless, these methods, can suffer too from reduced information
transfer rates~\cite{Chennu2013}.
Furthermore, these systems typically require users to focus on a central fixation point while
selecting peripheral targets.
This entails that they still rely to some extent on
eye motor control, often necessitating central gaze fixation.
\todo{advantages disadvantages rewrite (requires gaze fixation, low itr, not tested with split attention)}
\todo{
Firstly, by avoiding visual
stimulation entirely and opting for auditory or somatosensory stimulation instead.
However, these alternatives often result in lower information transfer rates,
increased mental effort and variable outcomes for different users.
Secondly, gaze-independence can
be achieved by adapting the interface to display stimuli such that they
are always in the center of the field of view, by exploiting visual
non-spatial attention (feature attention), or a combination thereof.
These interfaces suffer equally from reduced information transfer rates.
}

Thirdly, stimuli can be presented in a standard BCI paradigm, but visuospatial attention
can be decoded separately from gaze direction.
Earlier work has made advances in decoding lateralized covert
VSA harnessing the N2pc ERP component~\cite{Thiery2016,Reichert2020b,Wang2022},
decoding covert VSA shifts from spectral content~\cite{Tonin2013}, or hybrid
stimulation paradigms~\cite{Egan2017}.
Since these methods often require a slower stimulation pace than the classical
ERP BCI paradigm, gaze-independent decoding in fast-paced ERP paradigm stimulation
remains underexplored.
\cite{Aloise2012b} aimed to bridge the performance gap between covert and
overt VSA decoding performance.
They compared classical linear and non-linear ERP classifiers on a covert
attention P3 ERP component dataset.
The results revealed no significant performance improvement in covert VSA
decoding for any of the investigated decoders.
\cite{Arico2014} observed higher variability in single-trial P3 peak
latencies relative to stimulus onset during covert VSA compared to overt VSA.
This latency variability contributes to reduced covert VSA decoding performance.
While they proposed an analysis and performance prediction method, they did not provide a
decoding solution.
They suggested that compensating for latency jitter could enhance covert VSA
decoding, but did not actually verify this hypothesis directly.
Additionally, \cite{hardiansyah2020single} developed a classifier for
covert VSA ERPs, exploiting single-trial latency features in combination with
amplitude features for classification with a support vector machine.
They demonstrated the positive influence of single-trial ERP component latency
features on covert VSA inference yet did not attempt to correct the amplitude
features for these latencies.
\todo{Should we already mention Frenzel here?}
\cite{Frenzel2011} introduced a similar protocol to our split VSA
setting.
They showed that it is possible to perform split VSA and that, in this
case, visuospatial attention and gaze direction can separately be decoded using
classical ERP techniques.
To the best of our
knowledge, this is the sole study that investigated split attention in ERP-based BCIs.
However, \cite{Frenzel2011} considered their interface only for the case
where the user actively intends to select both targets determined by the gaze
and the VSA.
In the split VSA setting considered in our work, we rather instruct the participant to ignore
the distractor and only attend the cued target, since we are interested in
decoding the visuospatial attention only.
